---
title: 'Project 3 EDA'
author: We Came, We Saw, We Modelled | Emily Friedman (Group Leader), Jensen Harvey,
  Kaitlyn Chou
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    toc_depth: 2
    number_sections: yes
  html_document:
    toc: yes
    toc_depth: '2'
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.width  = 7,
  fig.height = 4.5
)
```


```{r}
library(jsonlite)
library(tidyverse)
library(ggplot2)
library(scales)
library(patchwork)
```

------------------------------------------------------------------------

# Data Loading

```{r}
data_dir <- "/Users/jensen.harvey/Desktop/data science/project3/data"

json_files <- list.files(
  path      = data_dir,
  pattern   = "gtBboxCityPersons\\.json$",
  recursive = TRUE,
  full.names = TRUE
)

cat("Total annotation files found:", length(json_files), "\n")

`%||%` <- function(a, b) if (!is.null(a)) a else b

parse_bbox_file <- function(filepath) {
  parts <- strsplit(filepath, "/")[[1]]
  n     <- length(parts)
  city  <- parts[n - 1]
  split <- parts[n - 2]

  ann <- tryCatch(fromJSON(filepath, simplifyVector = FALSE), error = function(e) NULL)
  if (is.null(ann)) return(NULL)

  img_w <- ann$imgWidth  %||% NA_integer_
  img_h <- ann$imgHeight %||% NA_integer_

  objects <- ann$objects
  if (is.null(objects) || length(objects) == 0) {
    return(data.frame(
      file       = filepath,
      split      = split,
      city       = city,
      img_width  = img_w,
      img_height = img_h,
      label      = NA_character_,
      bbox_x     = NA_real_,
      bbox_y     = NA_real_,
      bbox_w     = NA_real_,
      bbox_h     = NA_real_,
      bbox_vis_w = NA_real_,
      bbox_vis_h = NA_real_,
      stringsAsFactors = FALSE
    ))
  }

  rows <- lapply(objects, function(o) {
    bbox     <- o$bbox    %||% list(0, 0, 0, 0)
    bbox_vis <- o$bboxVis %||% list(0, 0, 0, 0)
    data.frame(
      file       = filepath,
      split      = split,
      city       = city,
      img_width  = img_w,
      img_height = img_h,
      label      = o$label %||% NA_character_,
      bbox_x     = bbox[[1]],
      bbox_y     = bbox[[2]],
      bbox_w     = bbox[[3]],
      bbox_h     = bbox[[4]],
      bbox_vis_w = bbox_vis[[3]],
      bbox_vis_h = bbox_vis[[4]],
      stringsAsFactors = FALSE
    )
  })
  do.call(rbind, rows)
}

annotations <- map_dfr(json_files, parse_bbox_file)

cat("Total annotated object rows:", nrow(annotations), "\n")
cat("Unique images:", n_distinct(annotations$file), "\n")
cat("Splits:", paste(sort(unique(annotations$split)), collapse = ", "), "\n")
cat("Cities:", paste(sort(unique(annotations$city)), collapse = ", "), "\n")
```

------------------------------------------------------------------------

# Dataset Overview

## Image Count by Split

```{r}
images_df <- annotations %>%
  distinct(file, split, city, img_width, img_height)

split_counts <- images_df %>%
  count(split, name = "n_images") %>%
  arrange(desc(n_images))

split_counts

ggplot(split_counts, aes(x = reorder(split, -n_images), y = n_images, fill = split)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = comma(n_images)), vjust = -0.4, size = 4) +
  scale_fill_manual(values = c("train" = "#2c7bb6", "val" = "#d7191c")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)), labels = comma) +
  labs(
    title = "Number of Images per Dataset Split",
    x     = "Split",
    y     = "Image Count"
  ) +
  theme_minimal(base_size = 12)
```

## Image Count by City

```{r}
city_counts <- images_df %>%
  count(city, split, name = "n_images")

ggplot(city_counts, aes(x = reorder(city, n_images), y = n_images, fill = split)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("train" = "#2c7bb6", "val" = "#d7191c")) +
  labs(
    title = "Image Count by City and Split",
    x     = "City",
    y     = "Number of Images",
    fill  = "Split"
  ) +
  theme_minimal(base_size = 11)
```

------------------------------------------------------------------------

# Label Distribution

```{r}
label_counts <- annotations %>%
  filter(!is.na(label)) %>%
  count(label, name = "n_objects") %>%
  arrange(desc(n_objects))

label_counts

ggplot(label_counts, aes(x = reorder(label, n_objects), y = n_objects, fill = label)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = comma(n_objects)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2)), labels = comma) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Annotation Label Counts Across All Images",
    x     = "Label",
    y     = "Number of Annotated Objects"
  ) +
  theme_minimal(base_size = 12)
```

## Person Annotations vs. Ignore Regions

```{r}
annotations <- annotations %>%
  mutate(
    is_ignore = grepl("ignore", label, ignore.case = TRUE),
    is_person = !is_ignore & !is.na(label)
  )

type_summary <- annotations %>%
  filter(!is.na(label)) %>%
  count(is_ignore) %>%
  mutate(
    type = ifelse(is_ignore, "Ignore Region", "Person Annotation"),
    pct  = n / sum(n)
  )

ggplot(type_summary, aes(x = type, y = n, fill = type)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = paste0(comma(n), "\n(", percent(pct, accuracy = 0.1), ")")),
            vjust = -0.3, size = 4) +
  scale_fill_manual(values = c("Person Annotation" = "#d73027",
                               "Ignore Region"     = "#878787")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18)), labels = comma) +
  labs(
    title = "Person Annotations vs. Ignore Regions",
    x     = NULL, y = "Count"
  ) +
  theme_minimal(base_size = 12)
```

------------------------------------------------------------------------

# Pedestrian Presence per Image (Binary Target Variable)

```{r}
image_level <- annotations %>%
  group_by(file, split, city, img_width, img_height) %>%
  summarise(
    n_persons        = sum(is_person,  na.rm = TRUE),
    n_ignore_regions = sum(is_ignore,  na.rm = TRUE),
    has_pedestrian   = any(is_person,  na.rm = TRUE),
    .groups = "drop"
  )

presence_summary <- image_level %>%
  count(has_pedestrian) %>%
  mutate(
    label = ifelse(has_pedestrian, "Has Pedestrian", "No Pedestrian"),
    pct   = n / sum(n)
  )

presence_summary

ggplot(presence_summary, aes(x = label, y = n, fill = label)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = paste0(comma(n), "\n(", percent(pct, accuracy = 0.1), ")")),
            vjust = -0.3, size = 4.2) +
  scale_fill_manual(values = c("Has Pedestrian" = "#d73027",
                               "No Pedestrian"  = "#4575b4")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18)), labels = comma) +
  labs(
    title    = "Image-Level Pedestrian Presence",
    subtitle = "Binary target variable for Faster R-CNN detection model",
    x = NULL, y = "Number of Images"
  ) +
  theme_minimal(base_size = 12)
```

## Presence Rate by Split

```{r}
split_presence <- image_level %>%
  group_by(split, has_pedestrian) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(label = ifelse(has_pedestrian, "Has Pedestrian", "No Pedestrian"))

ggplot(split_presence, aes(x = split, y = n, fill = label)) +
  geom_col(position = "fill", width = 0.5) +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = c("Has Pedestrian" = "#d73027",
                               "No Pedestrian"  = "#4575b4")) +
  labs(
    title = "Pedestrian Presence Rate by Split",
    x = "Split", y = "Proportion of Images", fill = NULL
  ) +
  theme_minimal(base_size = 12)
```

------------------------------------------------------------------------

# Pedestrian Count per Image

## Histogram

```{r}
image_level %>%
  filter(has_pedestrian) %>%
  ggplot(aes(x = n_persons)) +
  geom_histogram(binwidth = 1, fill = "#d73027", color = "white") +
  scale_x_continuous(breaks = pretty_breaks(n = 10)) +
  labs(
    title    = "Distribution of Person Count per Image",
    subtitle = "Among images containing at least one person annotation",
    x = "Number of Person Annotations", y = "Number of Images"
  ) +
  theme_minimal(base_size = 12)
```

## Summary Statistics

```{r}
image_level %>%
  summarise(
    total_images           = n(),
    images_with_pedestrian = sum(has_pedestrian),
    images_without         = sum(!has_pedestrian),
    pct_with_pedestrian    = mean(has_pedestrian),
    mean_persons_all       = mean(n_persons),
    mean_persons_ped_only  = mean(n_persons[has_pedestrian]),
    median_persons_ped     = median(n_persons[has_pedestrian]),
    max_persons            = max(n_persons)
  ) %>%
  pivot_longer(everything(), names_to = "Metric", values_to = "Value") %>%
  mutate(Value = round(Value, 3))
```

------------------------------------------------------------------------

# Bounding Box Analysis

```{r}
bbox_df <- annotations %>%
  filter(is_person, !is.na(bbox_w), !is.na(bbox_h), bbox_w > 0, bbox_h > 0) %>%
  mutate(
    bbox_area    = bbox_w * bbox_h,
    aspect_ratio = bbox_h / bbox_w,   
    vis_area     = bbox_vis_w * bbox_vis_h,
    visibility   = ifelse(bbox_area > 0, vis_area / bbox_area, NA_real_)
  )
```

## Width, Height, Area, and Aspect Ratio

```{r}
p1 <- ggplot(bbox_df, aes(x = bbox_w)) +
  geom_histogram(bins = 40, fill = "#4575b4", color = "white") +
  labs(title = "Bbox Width", x = "Width (px)", y = "Count") +
  theme_minimal(base_size = 10)

p2 <- ggplot(bbox_df, aes(x = bbox_h)) +
  geom_histogram(bins = 40, fill = "#74add1", color = "white") +
  labs(title = "Bbox Height", x = "Height (px)", y = "Count") +
  theme_minimal(base_size = 10)

p3 <- ggplot(bbox_df, aes(x = bbox_area)) +
  geom_histogram(bins = 40, fill = "#f46d43", color = "white") +
  scale_x_log10(labels = comma) +
  labs(title = "Bbox Area (log scale)", x = "Area (px²)", y = "Count") +
  theme_minimal(base_size = 10)

p4 <- ggplot(bbox_df, aes(x = aspect_ratio)) +
  geom_histogram(bins = 40, fill = "#fdae61", color = "white") +
  labs(title = "Aspect Ratio (H/W)", x = "H / W", y = "Count") +
  theme_minimal(base_size = 10)

(p1 + p2) / (p3 + p4)
```

## Visibility (Occlusion)

```{r}
bbox_df %>%
  filter(visibility >= 0, visibility <= 1) %>%
  ggplot(aes(x = visibility)) +
  geom_histogram(bins = 30, fill = "#1a9641", color = "white") +
  scale_x_continuous(labels = percent) +
  labs(
    title    = "Pedestrian Visibility Distribution",
    subtitle = "Fraction of bounding box that is unoccluded",
    x = "Visibility Fraction", y = "Count"
  ) +
  theme_minimal(base_size = 12)
```

## Bounding Box Summary Statistics

```{r}
bbox_df %>%
  summarise(
    n_person_annotations = n(),
    mean_width           = mean(bbox_w),
    median_width         = median(bbox_w),
    mean_height          = mean(bbox_h),
    median_height        = median(bbox_h),
    mean_area            = mean(bbox_area),
    median_area          = median(bbox_area),
    mean_aspect_ratio    = mean(aspect_ratio),
    mean_visibility      = mean(visibility, na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "Metric", values_to = "Value") %>%
  mutate(Value = round(Value, 2))
```

------------------------------------------------------------------------

# City-Level Analysis

## Pedestrian Presence Rate by City

```{r}
city_stats <- image_level %>%
  group_by(city, split) %>%
  summarise(
    n_images     = n(),
    n_with_ped   = sum(has_pedestrian),
    pct_with_ped = mean(has_pedestrian),
    avg_persons  = mean(n_persons),
    .groups      = "drop"
  )

ggplot(city_stats, aes(x = reorder(city, pct_with_ped),
                       y = pct_with_ped, fill = split)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = percent(pct_with_ped, accuracy = 1)),
            hjust = -0.1, size = 2.8) +
  coord_flip() +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.18))) +
  scale_fill_manual(values = c("train" = "#2c7bb6", "val" = "#d7191c")) +
  labs(
    title = "Pedestrian Presence Rate by City",
    x = "City", y = "% Images with >1 Person", fill = "Split"
  ) +
  theme_minimal(base_size = 11)
```

## Average Person Count by City

```{r}
ggplot(city_stats, aes(x = reorder(city, avg_persons), y = avg_persons, fill = split)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("train" = "#2c7bb6", "val" = "#d7191c")) +
  labs(
    title = "Average Person Annotations per Image, by City",
    x = "City", y = "Mean Person Count", fill = "Split"
  ) +
  theme_minimal(base_size = 11)
```

------------------------------------------------------------------------

# Class Imbalance Assessment

```{r}
imbalance <- image_level %>%
  count(has_pedestrian) %>%
  mutate(
    class = ifelse(has_pedestrian, "Positive (pedestrian)", "Negative (no pedestrian)"),
    pct   = n / sum(n)
  )

imbalance %>% select(class, n, pct) %>% mutate(pct = percent(pct, accuracy = 0.1))

ratio <- max(imbalance$n) / min(imbalance$n)
cat("Imbalance ratio (majority:minority):", round(ratio, 2), "\n")
cat(ifelse(ratio > 3,
  ">> Ratio > 3:1 — consider class-weighted loss or oversampling.",
  ">> Ratio <= 3:1 — class balance is acceptable for standard training."), "\n")
```

------------------------------------------------------------------------

# Image Dimensions

```{r}
dims_df <- images_df %>% filter(!is.na(img_width), !is.na(img_height))

pw <- ggplot(dims_df, aes(x = img_width)) +
  geom_histogram(bins = 20, fill = "#4575b4", color = "white") +
  labs(title = "Image Width Distribution", x = "Width (px)", y = "Count") +
  theme_minimal(base_size = 11)

ph <- ggplot(dims_df, aes(x = img_height)) +
  geom_histogram(bins = 20, fill = "#74add1", color = "white") +
  labs(title = "Image Height Distribution", x = "Height (px)", y = "Count") +
  theme_minimal(base_size = 11)

pw + ph
```

```{r}
dims_df %>%
  summarise(
    unique_resolutions = n_distinct(paste(img_width, img_height)),
    most_common_width  = as.integer(names(sort(table(img_width),  decreasing = TRUE)[1])),
    most_common_height = as.integer(names(sort(table(img_height), decreasing = TRUE)[1])),
    min_width = min(img_width),  max_width = max(img_width),
    min_height = min(img_height), max_height = max(img_height)
  )
```

------------------------------------------------------------------------

# Missing / Data Quality Check

```{r}
annotations %>%
  summarise(
    pct_missing_label    = mean(is.na(label)),
    pct_missing_bbox_w   = mean(is.na(bbox_w)),
    pct_missing_bbox_h   = mean(is.na(bbox_h)),
    pct_missing_img_dims = mean(is.na(img_width) | is.na(img_height))
  ) %>%
  pivot_longer(everything(), names_to = "Field", values_to = "Pct_Missing") %>%
  mutate(Pct_Missing = percent(Pct_Missing, accuracy = 0.01))
```

------------------------------------------------------------------------

# EDA Summary Table

```{r}
tibble(
  Metric = c(
    "Total annotation files (images)",
    "Train images",
    "Validation images",
    "Cities represented",
    "Total person annotations (non-ignore)",
    "Total ignore-region annotations",
    "Images with >1 pedestrian",
    "Images without pedestrian",
    "Pedestrian presence rate",
    "Mean persons per image (all)",
    "Mean persons per image (ped only)",
    "Class imbalance ratio"
  ),
  Value = c(
    comma(n_distinct(annotations$file)),
    comma(sum(images_df$split == "train")),
    comma(sum(images_df$split == "val")),
    as.character(n_distinct(annotations$city)),
    comma(sum(annotations$is_person, na.rm = TRUE)),
    comma(sum(annotations$is_ignore, na.rm = TRUE)),
    comma(sum(image_level$has_pedestrian)),
    comma(sum(!image_level$has_pedestrian)),
    percent(mean(image_level$has_pedestrian), accuracy = 0.1),
    as.character(round(mean(image_level$n_persons), 2)),
    as.character(round(mean(image_level$n_persons[image_level$has_pedestrian]), 2)),
    as.character(round(ratio, 2))
  )
)
```

------------------------------------------------------------------------

*Data: Cityscapes Dataset --- gtBboxCityPersons annotations (Cordts et
al., 2016)*
